---
title: "It's a bird, it's a plane - no it's a UFO!"
format: 
  dashboard:
    scrolling: true
author: Git Culture
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
knitr::opts_knit$set(root.dir = "/users/jacquelineferri/Desktop/sppw/finalprojrepo/")

library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(ggridges)
library(wordcloud)
library(tm)
library(stringr)
```

# NUFORC
```{r neon theme, echo=FALSE}
## Generating cool neon theme

neon_theme <- theme_minimal(base_size = 14) +
  theme(
    plot.background   = element_rect(fill = "black", color = NA),
    panel.background  = element_rect(fill = "black", color = NA),
    legend.background = element_rect(fill = "black", color = NA),
    legend.key        = element_rect(fill = "black", color = NA),
    text              = element_text(color = "white"),
    axis.text         = element_text(color = "white"),
    plot.title        = element_text(color = "white", face = "bold", size = 18),
    axis.title        = element_text(color = "white"),
    
    # Add neon gridlines
    panel.grid.major  = element_line(color = "#39FF14", linewidth = 0.3, linetype = "dashed"),
    panel.grid.minor  = element_line(color = "#39FF1450", linewidth = 0.1),
    
    # Optional: remove panel border for cleaner neon effect
    panel.border = element_blank()
  )

```

```{r file setup, echo=FALSE}
# Read the file
df <- read_csv("nuforc_county_weather_bigfoot.csv")

# Filter UFO data & create year-month
ufo_data <- df %>%
  filter(sighting_type == "UFO") %>%
  mutate(
    sighted_date = as.Date(sighted_date),
    year_month = floor_date(sighted_date, "month")
  )

# Top 5 shapes
top_shapes <- c("light", "circle", "triangle", "fireball", "sphere")

# We're looking at sightings per month 
monthly_total <- ufo_data %>%
  group_by(year_month) %>%
  summarise(monthly_count = n(), .groups = "drop") %>%
  arrange(year_month) %>%
  mutate(cumulative_total = cumsum(monthly_count))

# Top shapes
monthly_shapes <- ufo_data %>%
  filter(Data.Shape %in% top_shapes) %>%
  group_by(year_month, Data.Shape) %>%
  summarise(monthly_count = n(), .groups = "drop") %>%
  arrange(year_month, Data.Shape) %>%
  group_by(Data.Shape) %>%
  mutate(cumulative_count = cumsum(monthly_count)) %>%
  ungroup()
``` 

```{r plot builder, echo=FALSE}
#Cumulative sightings over time, neon theme
cumulative_sightings <- ggplot() +
  # Outer glow for total line
  geom_line(data = monthly_total,
            aes(x = year_month, y = cumulative_total),
            color = "#00FFFF", size = 3, alpha = 0.2) +
  # Core total line
  geom_line(data = monthly_total,
            aes(x = year_month, y = cumulative_total),
            color = "#00FFFF", size = 1) +
  
  # Shape-specific glowing lines
  geom_line(data = monthly_shapes,
            aes(x = year_month, y = cumulative_count, color = Data.Shape),
            size = 3, alpha = 0.15) +
  geom_line(data = monthly_shapes,
            aes(x = year_month, y = cumulative_count, color = Data.Shape),
            size = 1) +
  
  scale_color_manual(values = c(
    "light"    = "#00FFFF",
    "circle"   = "#FF00FF",
    "triangle" = "#39FF14",
    "fireball" = "#FF4500",
    "sphere"   = "#FFD700"
  )) +
  
  labs(
    title = "Cumulative UFO Sightings Over Time (Neon Style)",
    x = "Month",
    y = "Cumulative Sightings",
    color = "UFO Shape"
  ) +
  neon_theme

#CUMULATIVE SIGHTINGS
#-----------

#Cumulative sightings over time
UFO_shape_years <- ufo_data %>%
  ggplot(aes(x = year, fill = Data.Shape))+
  geom_bar()+
  theme_minimal()+
  labs(title="Shape over years?")+
  neon_theme

#TOP COUNTIES

#Neon theme hell yeah
top15_counties <- ufo_data %>%
  count(County, sort = TRUE) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(County, n), y = n)) +
  
  # Glow layer
  geom_col(fill = "#00FFFF", alpha = 0.25, width = 0.9) +
  # Core bar
  geom_col(fill = "#00FFFF", width = 0.6) +
  
  coord_flip() +
  labs(title = "Top 15 California Counties for UFO Sightings") +
  neon_theme

# SEASONAL PATTERNS
#------------------

## want to display months in box plots and use the pichart visual for the four seasons
### add column indicating season

# Create Seasons column
ufo_data <- ufo_data %>%
  mutate(Seasons = case_when(
    Dates.Sighted.Month %in% c(12, 1, 2) ~ "Winter",
    Dates.Sighted.Month %in% 3:5 ~ "Spring",
    Dates.Sighted.Month %in% 6:8 ~ "Summer",
    Dates.Sighted.Month %in% 9:11 ~ "Fall",
    TRUE ~ NA_character_  # in case of missing or invalid month
  ))
# create month name column
ufo_data <- ufo_data %>%
  mutate(Month.Name = case_when(
    Dates.Sighted.Month %in% 1 ~"Janaury",
    Dates.Sighted.Month %in% 2 ~ "February",
    Dates.Sighted.Month %in% 3 ~ "March",
    Dates.Sighted.Month %in% 4 ~ "April",
    Dates.Sighted.Month %in% 5 ~ "May",
    Dates.Sighted.Month %in% 6 ~ "June",
    Dates.Sighted.Month %in% 7 ~ "July",
    Dates.Sighted.Month %in% 8 ~ "August",
    Dates.Sighted.Month %in% 9 ~ "September",
    Dates.Sighted.Month %in% 10 ~ "October",
    Dates.Sighted.Month %in% 11 ~ "November",
    Dates.Sighted.Month %in% 12 ~ "December"
  ))

#Neon theme my man
seasonal_sightings <- ufo_data %>%
  count(Seasons) %>%
  ggplot(aes(x = as.factor(Seasons), y = n)) +
  geom_col(fill = "#FF00FF", alpha = 0.25, width = 1) +
  geom_col(fill = "#FF00FF", width = 0.7) +
  
  coord_polar() +
  labs(title = "Seasonal Pattern of UFO Sightings") +
  neon_theme

monthly_sightings <- ufo_data %>%
  group_by(Month.Name) %>%
  mutate(SightingNumber = row_number()) %>%
  ungroup() %>%
  ggplot(aes(x = as.factor(Month.Name), y = SightingNumber, fill = as.factor(Month.Name))) +
  geom_boxplot(color = "yellow") +  # bright yellow outline
  labs(x = "Month", y = "Number of sightings",title = "Count of Sightings by Month") +
  theme_minimal() +
  neon_theme +
  theme(
    legend.position = "none",  
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


#DISTRIBUTION PLOTS
#-----------------
#Neon boy
UFOsightings_time_space <- ufo_data %>%
  filter(sighting_type == "UFO") %>%
  ggplot(aes(x = sighted_date, y = Data.Shape, fill = Data.Shape)) +
  
  geom_density_ridges(alpha = 0.8, color = "black") +
  
  labs(title = "Distribution of UFO Sightings Over Time by Shape",
       x = "Date", y = "Shape") +
  neon_theme+
  theme(legend.position = "none")


#ENCOUNTER DURATION
#-----------------

#Neon
UFOencounter_duration_shape <- ufo_data %>%
  filter(sighting_type == "UFO") %>%
  group_by(Data.Shape) %>%
  summarise(mean_duration = mean(as.numeric(Data.Encounter.duration), na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(Data.Shape, mean_duration), y = mean_duration)) +
  
  geom_col(fill = "#39FF14", alpha = 0.3) +
  geom_col(fill = "#39FF14", width = 0.6) +
  
  coord_flip() +
  labs(title = "Average Encounter Duration by UFO Shape",
       x = "Shape", y = "Mean Duration (seconds)") +
  neon_theme

#WORDCLOUD
# Load required packages
library(tm)
library(wordcloud)

# Define the function to return a plot object
create_neon_wordcloud <- function(data_column, max_words = 100) {
  # Prepare the text
  text <- Corpus(VectorSource(data_column))
  text <- tm_map(text, content_transformer(tolower))
  text <- tm_map(text, removePunctuation)
  text <- tm_map(text, removeNumbers)
  text <- tm_map(text, removeWords, stopwords("english"))
  
  # Define neon colors
  neon_colors <- c("#00FFFF", "#FF00FF", "#39FF14", "#FFD700", "#FF4500", "#00FF7F")
  
  # Open off-screen device
  tmp <- tempfile(fileext = ".png")
  png(tmp, width = 800, height = 600, bg = "black")
  
  # Draw wordcloud
  wordcloud(
    words = text,
    max.words = max_words,
    scale = c(3, 0.5),
    random.order = FALSE,
    rot.per = 0.25,
    colors = neon_colors
  )
  
  # Capture the plot
  wc_plot <- recordPlot()
  
  # Close device
  dev.off()
  
  return(wc_plot)
}

# Usage
plot <- create_neon_wordcloud(ufo_data$Data.Description.excerpt, max_words = 100)

# Display later
replayPlot(plot)

## Temperature
#-----------

#Neon
UFOsightings_temperature <- ufo_data %>%
  group_by(Avg_Temp_Month_County) %>%
  summarise(count = n()) %>%
  ggplot(aes(Avg_Temp_Month_County, count)) +
  
  geom_point(color = "#FFD700", size = 3, alpha = 0.9) +
  geom_smooth(color = "#FF00FF", size = 1) +
  
  labs(title = "Do UFO Sightings Correlate with Temperature?",
       x = "Average Monthly Temperature (°F)",
       y = "Sightings") +
  neon_theme


#DAY VS NIGHT
#------------
# NEON
day_night_UFOsightings <- ufo_data %>%
  mutate(time_of_day = ifelse(Dates.Sighted.Hour >= 6 & Dates.Sighted.Hour < 18, "Day", "Night")) %>%
  count(time_of_day) %>%
  ggplot(aes(time_of_day, n, fill = time_of_day)) +
  
  geom_col(alpha = 0.25) +
  geom_col(width = 0.6) +
  
  scale_fill_manual(values = c(
    "Day" = "#00FFFF",
    "Night" = "#FF00FF"
  )) +
  
  labs(title = "Day vs Night UFO Sightings",
       x = "", y = "Sightings") +
  neon_theme
```
## 
issue: the wordcloud won't print out 2025.12.06 JTF
```{r, echo=FALSE}
# Display the stored plot later
replayPlot(plot)

```

## {width=20}
```{r, echo=FALSE}
cumulative_sightings
```
## {width=20}
```{r, echo=FALSE}
day_night_UFOsightings

```
## {width=20}
```{r, echo=FALSE}
UFOsightings_temperature

```
## {width=20}

```{r, echo=FALSE}
monthly_sightings
```
## {width=20}

```{r, echo=FALSE}
seasonal_sightings

```
## {width=20}
```{r, echo=FALSE}
UFOsightings_time_space
```
## {width=20}
```{r, echo=FALSE}
top15_counties
```


```{r, echo=FALSE}
UFOencounter_duration_shape

```

# Health data

## YPLL
```{r, echo=FALSE}
caliYPLL <- read.csv("county_health_data/combined_california_ypll.csv")
nuforc <- read.csv("nuforc_county_weather_bigfoot.csv")

## can use ggplot2 and have mutliple separate layers per plot and pull from different dataframes
library(ggplot2)
library(tigris)
library(dplyr)
library(ggplot2)
library(sf)
library(viridis)

options(tigris_use_cache = TRUE)

# 1. Load US county shapefiles
counties_sf <- counties(cb = TRUE, year = 2023)  # cb = cartographic boundaries

# 2. Make sure FIPS codes match format (zero-padded 5-digit strings)
caliYPLL$FIPS <- sprintf("%05d", as.numeric(caliYPLL$FIPS))

# 3. Join your data to the map data using FIPS
map_df <- counties_sf %>%
  mutate(FIPS = GEOID) %>%
  left_join(caliYPLL, by = "FIPS")

# 4. Plot choropleth
YPLL_plot <- ggplot(map_df %>% filter(STATEFP == "06")) +
  geom_sf(aes(fill = YPLL_Rate), color = NA) +     # NA removes borders for cleaner heat map
  scale_fill_viridis(option = "magma", direction = -1, na.value = "grey90") +
  labs(
    title = "Years of Potential Life Lost Rate by County",
    fill = "YPLL Rate"
  ) +
  theme_minimal()

## format ufo_data for same visualization
## clean data so names match
ufo_data <- ufo_data %>%
  mutate(
    County = str_to_title(County),     # "ALAMEDA" → "Alameda"
    County = paste0(County, " County"),
    State = str_to_title(Location.State)        # if you have a State column
  )

## prepare shapefile names
counties_sf <- counties(cb = TRUE, year = 2023) %>%
  mutate(
    County = paste0(NAME, " County"),
    State = str_to_title(STATE_NAME)
  )

ufo_map_df <- counties_sf %>%
  left_join(ufo_data, by = c("State", "County"))

# Reproject shapefile (fixes warning #1)
counties_sf_proj <- st_transform(counties_sf, 5070)

# plot
ggplot(
  counties_sf_proj %>%
    filter(STUSPS == "CA") %>%   # only California
    left_join(
      ufo_data %>%
        filter(State == "CA") %>%
        count(State, County, name = "sightings"),   # count inline
      by = c("STUSPS" = "State", "NAME" = "County")
    )
) +
  geom_sf(aes(fill = sightings), color = "white", size = 0.2) +
  scale_fill_viridis_c(
    option = "magma",
    na.value = "grey90",
    name = "Sightings"
  ) +
  labs(
    title = "UFO Sightings per County in California",
    subtitle = "Number of sightings per county"
  ) +
  theme_minimal()

sightings_heatmap <- ggplot(ufo_map_df %>% filter(STATEFP == "06"))+
  geom_sf(aes(fill="County"), color = NA)+
  scale_fill_viridis(option="magma", direction = -1, na.value = "grey90")+
  labs(
    title = "Sightings by county",
    fill = "County")+
  theme_minimal()
  )
YPLL_plot <- ggplot(map_df %>% filter(STATEFP == "06")) +
  geom_sf(aes(fill = YPLL_Rate), color = NA) +     # NA removes borders for cleaner heat map
  scale_fill_viridis(option = "magma", direction = -1, na.value = "grey90") +
  labs(
    title = "Years of Potential Life Lost Rate by County",
    fill = "YPLL Rate"
  ) +
  theme_minimal()
```

```{r}
# function to create nuforc variable heatmaps
library(dplyr)
library(purrr)
options(tigris_use_cache = FALSE)  # optional, avoids caching errors

# Columns/variables you want
variables_to_count <- c(
  "City",
  "County",
  "Data.Shape",
  "Seasons",
  "Dates.Sighted.Year",
  "Date.Sighted.Day",
  "Dates.Sighted.Month"
)

# counties_sf already loaded with CA counties
counties_clean <- counties_sf

library(dplyr)
library(ggplot2)
library(purrr)
library(stringr)
library(sf)
library(viridis)

# Assume counties_sf already loaded with CA counties
counties_clean <- counties_sf_proj

plot_auto_heatmaps <- function(counts_list, show_labels = FALSE) {
  
  plots <- map(names(counts_list), function(var_name) {
    
    df <- counts_list[[var_name]] %>%
      rename(value = 1, count = 2) %>%
      mutate(value = as.character(value))
    
    # Detect if this is a county column
    is_county <- any(str_detect(df$value, "County$")) || any(tolower(df$value) %in% tolower(counties_clean$NAME))
    
    if(is_county) {
      # Clean county names by removing " County"
      df <- df %>% mutate(value = str_remove(value, " County$"))
      
      # Join with CA counties shapefile
      map_df <- counties_clean %>%
        left_join(df, by = c("NAME" = "value")) %>%
        mutate(count = ifelse(is.na(count), 0, count))
      
      p <- ggplot(map_df) +
        geom_sf(aes(fill = count), color = "white", size = 0.2)
      
      if(show_labels) {
        p <- p + geom_sf_text(aes(label = count), size = 3, color = "black")
      }
      
      p <- p +
        scale_fill_viridis_c(option = "magma", na.value = "grey90") +
        labs(
          title = paste("UFO Heatmap for", var_name, "- California"),
          fill = "Count"
        ) +
        theme_minimal()
      
      return(p)
      
    } else {
      # For non-county variables, make a bar plot
      df <- df %>% arrange(desc(count))
      p <- ggplot(df, aes(x = reorder(value, -count), y = count, fill = count)) +
        geom_col() +
        scale_fill_viridis_c(option = "magma") +
        labs(
          title = paste("UFO Sightings by", var_name),
          x = var_name,
          y = "Count",
          fill = "Count"
        ) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
      
      return(p)
    }
  })
  
  names(plots) <- names(counts_list)
  return(plots)
}

# Example usage
heatmaps <- plot_auto_heatmaps(ufo_data_counts, show_labels = TRUE)





