---
author: "Ananya Bhaktaram"
title: "NUFORC Data Scraping"
date: today
format: html
editor: visual
---

Load working libraries

```{r}
library(rvest)
library(dplyr)
library(readr)
library(httr)
library(stringr)
```

### Analyze NUFORC Page Structure

```{r}
analyze_nuforc_page <- function(url) {
  
  # Fetch the page
  page <- read_html(url)
  
  # Check for tables
  tables <- html_nodes(page, "table")
  cat("Number of tables found:", length(tables), "\n")
  
  if (length(tables) > 0) {
    # Analyze first table
    first_table <- tables[[1]]
    
    # Check table attributes
    table_attrs <- html_attrs(first_table)
    cat("Table attributes:", paste(names(table_attrs), table_attrs, sep="=", collapse=", "), "\n")
    
    # Count rows and columns
    rows <- html_nodes(first_table, "tr")
    cat("Number of rows in first table:", length(rows), "\n")
    
    if (length(rows) > 0) {
      first_row_cells <- html_nodes(rows[[1]], "td, th")
      cat("Number of columns:", length(first_row_cells), "\n")
      
      # Show first few cell contents
      cat("\nFirst row contents:\n")
      for (i in seq_along(first_row_cells)[1:min(5, length(first_row_cells))]) {
        cell_text <- html_text(first_row_cells[[i]], trim = TRUE)
        cat("  Column", i, ":", substr(cell_text, 1, 50), "\n")
      }
    }
  }
  
  # Check for JavaScript/AJAX indicators
  scripts <- html_nodes(page, "script")
  cat("\nNumber of script tags:", length(scripts), "\n")
  
  # Look for common AJAX/dynamic loading indicators
  ajax_indicators <- c("ajax", "xhr", "fetch", "load", "dynamic")
  script_content <- paste(html_text(scripts), collapse = " ")
  found_indicators <- ajax_indicators[sapply(ajax_indicators, function(x) grepl(x, script_content, ignore.case = TRUE))]
  
  if (length(found_indicators) > 0) {
    cat("Possible AJAX/dynamic loading indicators found:", paste(found_indicators, collapse = ", "), "\n")
  }
  
  # Check for pagination
  pagination_selectors <- c("a[href*='page']", ".pagination", ".pager", "a[href*='next']", "a[href*='prev']")
  pagination_found <- FALSE
  
  for (selector in pagination_selectors) {
    elements <- html_nodes(page, selector)
    if (length(elements) > 0) {
      cat("Pagination elements found with selector '", selector, "':", length(elements), "\n")
      pagination_found <- TRUE
    }
  }
  
  if (!pagination_found) {
    cat("No obvious pagination elements found\n")
  }
  
  # Check page size and load time
  page_content <- as.character(page)
  cat("\nPage size: ~", round(nchar(page_content)/1024), "KB\n")
  
  return(invisible(page))
}

# Run analysis
url <- "https://nuforc.org/subndx/?id=all"
analyze_nuforc_page(url)
```

### Test Scrape

```{r}
library(rvest)
library(dplyr)

# Test Scrape
test_scrape <- function() {
  url <- "https://nuforc.org/subndx/?id=all"
  
  # Try with a longer delay 
  Sys.sleep(5)  
  page <- read_html(url)
  
  # Look for tables
  tables <- html_table(page, fill = TRUE)
  cat("Found", length(tables), "tables\n")
  
  if (length(tables) > 0) {
    for (i in 1:length(tables)) {
      df <- tables[[i]]
      cat("Table", i, ":", nrow(df), "rows x", ncol(df), "columns\n")
      
      if (nrow(df) > 1) {
        cat("First few rows of table", i, ":\n")
        print(head(df, 3))
        cat("\n")
      }
    }
    
    return(tables)
  } else {
    return(NULL)
  }
}

test_data <-test_scrape()
```

Using rvest worked to extract the first 100 entries. Explore table structure and layout of the sample table

The columns we want to extract are:

Occurred, City, State, Country, Shape, and Reported.

We want to restrict data only to observed sightings in the United States

### Clean Test Data

```{r}
# Clean Extracted Data
process_ufo <- function(test_data) {

  # Load table
  df <- test_data[[1]]
  
   # Clean data
  clean_df <- df %>%
    # Format text columns
    mutate(across(everything(), ~ {
      text <- as.character(.)
      text <- str_trim(str_squish(text))  # Remove extra whitespace
      text[text == "" | text == " " | text == "NULL"] <- NA
      return(text)
    })) %>%
    # Remove any completely emptry rows
    filter(if_any(everything(), ~ !is.na(.) & . != ""))
  
  # US state abbreviations for validation
  us_states <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "DC")
  
  # Filter for US entries 
  us_data <- clean_df %>%
    filter(
      # State is a US state OR Country contains USA/US
      State %in% us_states | 
      str_detect(toupper(Country), "USA|US|UNITED STATES")
    ) %>%
    # Select only the columns you requested
    select(Occurred, City, State, Country, Shape, Reported) %>%
    # Remove any rows where all selected columns are empty
    filter(if_any(everything(), ~ !is.na(.) & . != "")) %>%
    # Remove duplicates
    distinct()
  
  return(us_data)
}

# Process the data
clean_test <- process_ufo(test_data)
```

### Save Test Data

```{r}
# Save
save<- function(clean_test, output_file = "scraped_nuforc.csv") {
  
  
  # Save to CSV
  write_csv(clean_test, output_file)
  
  # Show sample
  print(head(clean_test, 5))
  
  return(clean_test)
}

# Save the data
clean_test_data <- save(clean_test)
```

### Use RSelenium Package to Extract all Pages

```{r}
library(rvest)
library(dplyr)
library(stringr)
library(readr)
library(httr)

# Enhanced rvest approach with session management
full_extraction <- function() {
  url <- "https://nuforc.org/subndx/?id=all"
  
  # Create a persistent session with realistic headers
  sess <- session(url) %>%
    session_jump_to(
      url,
      user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36 (KHTML, like Gecko) Chrome/120.0.0.0 Safari/537.36"),
      add_headers(
        "Accept" = "text/html,application/xhtml+xml,application/xml;q=0.9,image/webp,*/*;q=0.8",
        "Accept-Language" = "en-US,en;q=0.5",
        "Accept-Encoding" = "gzip, deflate, br",
        "Connection" = "keep-alive",
        "Upgrade-Insecure-Requests" = "1"
      )
    )
  
  # Try multiple approaches with increasing wait times
  best_result <- NULL
  max_rows <- 0
  
  for (attempt in 1:5) {
    cat("Attempt", attempt, "- waiting", attempt * 5, "seconds for JavaScript...\n")
    
    # Wait for JavaScript to load
    Sys.sleep(attempt * 5)
    
    # Re-read the page 
    page <- read_html(sess)
    
    # Extract tables
    tables <- html_table(page, fill = TRUE)
    
    if (length(tables) > 0) {
      df <- tables[[1]]
      
      if (nrow(df) > max_rows) {
        max_rows <- nrow(df)
        best_result <- df
      }
      
      # Check if there is greater than a 1000 entries
      if (nrow(df) > 1000) {
        break
      }
    }
    
    # Refresh the session to grab more data
    if (attempt < 5) {
      sess <- session_jump_to(sess, url)
    }
  }
  
  if (!is.null(best_result)) {
    cat("✅ Best result:", nrow(best_result), "rows\n")
    return(best_result)
  } else {
    cat("❌ No data found\n")
    return(NULL)
  }
}

# Run the enhanced rvest approach
nuforc_data2 <- full_extraction()
```

### Attempt 1 to fix encoding error

```{r}
library(httr)
library(rvest)

url <- "https://nuforc.org/subndx/?id=all"

# Modify basic headers
response <- GET(url, 
                user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36"),
                add_headers("Accept-Encoding" = "gzip, deflate"))

page <- read_html(content(response, "text"))
tables <- html_table(page, fill = TRUE)

```

### Testing URL Parameters to scrape more than the first 100 entries

```{r}
library(httr)
library(rvest)
library(dplyr)
library(stringr)
library(readr)

# Build on the working quick fix to get more data
more_data <- function() {
  
    base_url <- "https://nuforc.org/subndx/?id=all"
  
  # Test different URL parameters that might give us more data
  url_variations <- c(
    base_url,
    paste0(base_url, "&length=1000"),
    paste0(base_url, "&length=-1"),
    paste0(base_url, "&show=all"),
    paste0(base_url, "&limit=10000"),
    paste0(base_url, "&per_page=1000"),
    paste0(base_url, "&pageSize=1000"),
    paste0(base_url, "&table_1_length=1000"),
    paste0(base_url, "&table_1_length=-1"),
    paste0(base_url, "&draw=1&start=0&length=10000"),
    paste0(base_url, "&draw=1&start=0&length=-1")
  )
  
  best_result <- NULL
  max_rows <- 101  # We know we can get at least 101
  
  for (i in seq_along(url_variations)) {
    url <- url_variations[[i]]
    cat("Trying variation", i, ":", substr(url, 50, nchar(url)), "\n")
    
    tryCatch({
      # Use the same approach that worked
      response <- GET(url, 
                      user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36"),
                      add_headers("Accept-Encoding" = "gzip, deflate"),
                      timeout(60))
      
      if (status_code(response) == 200) {
        page <- read_html(content(response, "text"))
        tables <- html_table(page, fill = TRUE)
        
        if (length(tables) > 0) {
          df <- tables[[1]]
          cat("  Result:", nrow(df), "rows\n")
          
          if (nrow(df) > max_rows) {
            max_rows <- nrow(df)
            best_result <- df
            cat("  ✅ NEW BEST RESULT:", nrow(df), "rows!\n")
          }
        } else {
          cat("  No tables found\n")
        }
      } else {
        cat("  HTTP Status:", status_code(response), "\n")
      }
      
      # Pause between request processes to allow for system reload
      Sys.sleep(2)
      
    }, error = function(e) {
      cat("  Error:", conditionMessage(e), "\n")
    })
  }
  
  if (!is.null(best_result)) {
    cat("\n✅ Best result:", nrow(best_result), "rows\n")
    return(best_result)
  } else {
    cat("\n❌ No improvement over 101 rows\n")
    return(NULL)
  }
}

# Try URL variations
nuforc_data3 <- more_data()
```

### Testing Pagination

```{r}
library(httr)
library(rvest)
library(dplyr)
library(stringr)

# Attempted Pagination Function
scrape_multiple_pages <- function(max_pages = 20) {
  
  base_url <- "https://nuforc.org/subndx/?id=all"
  all_data <- data.frame()
  
  # Try different pagination patterns
  pagination_patterns <- list(
    # Standard pagination
    function(page) paste0(base_url, "&page=", page),
    
    # Offset-based
    function(page) paste0(base_url, "&start=", (page-1) * 100),
    function(page) paste0(base_url, "&offset=", (page-1) * 100),
    
    # DataTables style
    function(page) paste0(base_url, "&draw=", page, "&start=", (page-1) * 100, "&length=100"),
    
    # WordPress style
    function(page) paste0(base_url, "&paged=", page),
    
    # Other common patterns
    function(page) paste0(base_url, "&p=", page),
    function(page) paste0(base_url, "&pagenum=", page)
  )
  
  for (pattern_num in seq_along(pagination_patterns)) {
    cat("Trying pagination pattern", pattern_num, "...\n")
    
    pattern_data <- data.frame()
    consecutive_failures <- 0
    
    for (page in 1:max_pages) {
      url <- pagination_patterns[[pattern_num]](page)
      cat("  Page", page, "...")
      
      tryCatch({
        # Use our working method with the encoding fix
        response <- GET(url, 
                        user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36"), # Assuming Google Chrome Usage 
                        add_headers("Accept-Encoding" = "gzip, deflate"),
                        timeout(30))
        
        if (status_code(response) == 200) {
          page_html <- read_html(content(response, "text"))
          tables <- html_table(page_html, fill = TRUE)
          
          if (length(tables) > 0 && nrow(tables[[1]]) > 1) {
            page_data <- tables[[1]]
            
            # Check if this is new data (different from previous pages)
            if (nrow(pattern_data) == 0) {
              # First page for this pattern
              pattern_data <- page_data
              consecutive_failures <- 0
              cat(" ✅", nrow(page_data), "rows (Total:", nrow(pattern_data), ")\n")
            } else {
              # Check if data is different from what we already have
              # Compare first few rows to detect duplicates
              first_few_new <- page_data[1:min(3, nrow(page_data)), 1:min(3, ncol(page_data))]
              first_few_existing <- pattern_data[1:min(3, nrow(pattern_data)), 1:min(3, ncol(pattern_data))]
              
              if (!identical(first_few_new, first_few_existing)) {
                # New data found
                pattern_data <- bind_rows(pattern_data, page_data)
                consecutive_failures <- 0
                cat(" ✅", nrow(page_data), "new rows (Total:", nrow(pattern_data), ")\n")
              } else {
                consecutive_failures <- consecutive_failures + 1
                cat(" ⚠️ Duplicate data\n")
                
                if (consecutive_failures >= 3) {
                  cat("  Stopping pattern", pattern_num, "- too many duplicates\n")
                  break
                }
              }
            }
          } else {
            consecutive_failures <- consecutive_failures + 1
            cat(" ❌ No data\n")
            
            if (consecutive_failures >= 3) {
              cat("  Stopping pattern", pattern_num, "- no more data\n")
              break
            }
          }
        } else {
          cat(" HTTP", status_code(response), "\n")
          consecutive_failures <- consecutive_failures + 1
        }
        
        # Be respectful - wait between requests
        Sys.sleep(2)
        
      }, error = function(e) {
        cat(" Error:", conditionMessage(e), "\n")
        consecutive_failures <- consecutive_failures + 1
      })
      
      if (consecutive_failures >= 5) {
        cat("  Too many failures, stopping pattern", pattern_num, "\n")
        break
      }
    }
    
    if (nrow(pattern_data) > nrow(all_data)) {
      all_data <- pattern_data
      cat("✅ Pattern", pattern_num, "is best so far:", nrow(all_data), "total rows\n")
    } else {
      cat("Pattern", pattern_num, "result:", nrow(pattern_data), "rows\n")
    }
    
    cat("\n")
  }
  
  if (nrow(all_data) > 101) {
    cat("🎉 SUCCESS! Found", nrow(all_data), "total records across multiple pages!\n")
    return(all_data)
  } else {
    cat("❌ No pagination pattern found more than 101 records\n")
    return(NULL)
  }
}

# Now run the pagination function
multi_page_data <- scrape_multiple_pages(max_pages = 15)
```

## Latest (11/29/2025) Scrape Attempt - more manual less automated

### HTML Examination

```{r}
library(httr)    
library(rvest)   
library(dplyr)   
library(stringr) 
library(readr) 

# Load NUFORC website into R
url <- "https://nuforc.org/subndx/?id=all"

# Get the page (using the working method from above)
response <- GET(url, 
                user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36"), # Chrome user agent specification 
                add_headers("Accept-Encoding" = "gzip, deflate"))

page <- read_html(content(response, "text"))

# View the HTML Structure for the first 3000 characters
cat("=== NUFORC HTML STRUCTURE ===\n")
page %>%
  as.character() %>%
  substr(1, 3000) %>%
  cat()
```

The website appears to be in HTML, but the page is loading using JavaScript.

### Exploration for specific table structure

```{r}
# Find all tables
cat("Number of tables:", page %>% html_elements("table") %>% length(), "\n")

# Look for table classes and IDs
tables <- page %>% html_elements("table")
if (length(tables) > 0) {
  table_attrs <- html_attrs(tables[[1]])
  cat("First table attributes:\n")
  print(table_attrs)
}

# Find table headers
cat("\nTable headers:\n")
headers <- page %>% html_elements("table th, table tr:first-child td") %>% html_text()
print(headers)

# Find table rows
cat("\nNumber of table rows:", page %>% html_elements("table tr") %>% length(), "\n")

# Look at first few data rows
rows <- page %>% html_elements("table tr")
for (i in 1:min(5, length(rows))) {
  cells <- rows[[i]] %>% html_elements("td, th") %>% html_text(trim = TRUE)
  if (length(cells) > 0) {
    cat("Row", i, ":", paste(cells[1:min(5, length(cells))], collapse = " | "), "\n")
  }
}

# Look for DataTables specific elements
cat("Elements with 'dataTables' class:", page %>% html_elements("[class*='dataTables']") %>% length(), "\n")
cat("Elements with 'wpDataTable' class:", page %>% html_elements("[class*='wpDataTable']") %>% length(), "\n")
cat("Table with ID 'table_1':", page %>% html_elements("#table_1") %>% length(), "\n")

# Look for select dropdowns
selects <- page %>% html_elements("select")
cat("Number of select elements:", length(selects), "\n")

if (length(selects) > 0) {
  for (i in seq_along(selects)) {
    select_name <- selects[[i]] %>% html_attr("name")
    select_id <- selects[[i]] %>% html_attr("id") 
    options <- selects[[i]] %>% html_elements("option") %>% html_text()
    
    cat("Select", i, "- Name:", select_name %||% "none", 
        "ID:", select_id %||% "none", "\n")
    cat("  Options:", paste(options, collapse = ", "), "\n")
  }
}
```

There is 1 table, the table is being stored in HTML as a WordPress DataTable Plugin. There aren't an select elements, and the data is stored in the HTML rather than on a different server somewhere and simply being called to display.

The reason I have only been able to extract a 100 entries is because the table is hidden by CSS given that the display function is listed as none, and is being moderated by the JavaScript loading, and I have only been able to extract the first 100 records because that's what is defaultly loaded.

### Exploring Javascript Pagination

```{r}
library(rvest)
library(dplyr)
library(stringr)
library(readr)
library(chromote)


# Start browser
b <- chromote::default_chromote_object()$new_session()
b$Page$navigate("https://nuforc.org/subndx/?id=all")
Sys.sleep(30)

#Confrim page has loaded
page_check_js <- '
{
  title: document.title,
  url: window.location.href,
  bodyText: document.body ? document.body.innerText.substring(0, 500) : "no body",
  hasTable: document.querySelectorAll("table").length,
  hasDiv: document.querySelectorAll("div").length,
  scripts: document.querySelectorAll("script").length,
  readyState: document.readyState
}
'

page_info <- b$Runtime$evaluate(page_check_js)
page_data <- page_info$result$value

# Function for page discovery 
if (page_data$hasTable > 0){
  pagination_discovery_js <- '
  const results = {
    tables: [],
    pagination_containers: [],
    buttons: [],
    links: [],
    selects: []
  };

  const tables = document.querySelectorAll("table");
  tables.forEach((table, i) => {
    results.tables.push({
      index: i,
      id: table.id || "no-id",
      class: table.className || "no-class", 
      rows: table.querySelectorAll("tr").length,
      visible: table.offsetParent !== null
    });
  });

  const paginationSelectors = [
    ".dataTables_paginate", ".pagination", ".page-numbers",
    "[class*=paginate]", "[class*=paging]", ".wp-pagenavi"
  ];

  paginationSelectors.forEach(selector => {
    const elements = document.querySelectorAll(selector);
    if (elements.length > 0) {
      results.pagination_containers.push({
        selector: selector,
        count: elements.length,
        visible: elements[0].offsetParent !== null
      });
    }
  });

  const buttons = document.querySelectorAll("button, input[type=button]");
  buttons.forEach((btn, i) => {
    const text = btn.textContent.trim().toLowerCase();
    if (text.includes("next") || text.includes("prev") || text.match(/^\\\\d+$/)) {
      results.buttons.push({
        text: btn.textContent.trim(),
        visible: btn.offsetParent !== null
      });
    }
  });

  const links = document.querySelectorAll("a");
  links.forEach((link, i) => {
    const text = link.textContent.trim().toLowerCase();
    if (text.includes("next") || text.includes("prev") || text.match(/^\\\\d+$/)) {
      results.links.push({
        text: link.textContent.trim(),
        visible: link.offsetParent !== null
      });
    }
  });

  const selects = document.querySelectorAll("select");
  results.selects = Array.from(selects).map(select => ({
    name: select.name || "no-name",
    visible: select.offsetParent !== null,
    optionCount: select.options.length
  }));

  results;
  '
  
  discovery_result <- b$Runtime$evaluate(pagination_discovery_js)
  discovery_data <- discovery_result$result$value
  
}

# Close browser
b$close()
```

### 
